---
title: "EDLD 652 Lab 3"
author: "Ksenia Gordeeva, Rebeccca Gordon, Amy Warnock"
date: "1/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rio)
library(here)
```

### 1. Visualize the total production of honey across years by state Use color to highlight the west coast (Washington, Oregon, and California) with a different color used for each west coast state.

```{r part1}

```

### 2. Reproduce the plot according three different kinds of color blindness, as well as a desaturated version.

```{r part2}

```

### 3. Reproduce the plot using a color blind safe palette.

```{r part3}

```

### 4. Download the file here denoting the region and division of each state.  
* Join the file with your honey file.
* Produce a bar plot displaying the average honey for each state (collapsing across years).
* Use color to highlight the region of the country the state is from.
* Note patterns you notice.

```{r part4}

```

### 5. Create a heatmap displaying the average honey production across years by region (averaging across states within region).

```{r part5}

```

### 6. Create at least one more plot of your choosing using color to distinguish, represent data values, or highlight. 

```{r part6}

```


